Scaffolder - Software for Manual Genome Scaffolding.

### 2011-09-12 src/article.tex
Proof read again

The sequencing process generates reads from either strand of the DNA helix and
the resulting contigs constructed from these reads may represent either the
5'~$\rightarrow$~3' or 3'~$\rightarrow$~5' DNA strands with respect to the
origin of replication. Orientating all contigs to point in the same direction
requires reverse complementing sequences where necessary. In the case of
archaeal and bacterial genomes this orientation will be to the
5'~$\rightarrow$~3' direction following the direction of genome replication.

@@

During the scaffolding process, closing and filling gaps between contigs
completes and improves the genome scaffold. Closing gaps may require returning
to the laboratory to perform additional sequencing or using computational
methods of estimating the unknown sequence. This additional sequence is used
to replace the gap between two contigs, joining them into a single sequence.
Once all contigs have been joined and gaps in a scaffold closed, the genome
may be considered finished.

@@

The process of finishing a genome scaffold uses wet laboratory methods,
\emph{in silico} methods, or a combination of both. An example of a
computational method might use the paired-read data from the sequencing stage.
The occurrence of paired reads in separate contigs can be used to estimate
probabilistically the order and distance between these contigs. Alternatively,
laboratory methods may use PCR to amplify the unknown DNA in a gap region then
use Sanger sequencing to determine the sequence of this gap. Computational
methods using the available sequencing data are preferable as they are less
costly in laboratory time and materials compared to manual gap resolution
\cite{nagarajan2010}. Finally when the scaffold cannot be completely resolved,
\emph{in silico} software packages exist to suggest the likely primers
necessary for PCR amplifying the sequence in gap regions \cite{gordon2001}. \pb

@@

Initial sequencing data assembly may result in an incomplete genome build.
Generation of additional sequences either through PCR or computational methods
means that genome scaffolding may be an on-going process. The scaffold file
should therefore be simple to update manually in addition to being
computationally tractable. This requirement was also best suited to YAML syntax
which is human-readable and simple to edit in a standard text editor. \pb

@@

The input data for Scaffolder is nucleotide sequence in FASTA format file.
These nucleotide sequences can be of any length and for example may be
individual reads, assembled contigs or contigs which have been joined to into
larger scaffolds. The case in which Scaffolder may be most useful is using the
contigs and scaffolded contigs, combined with additional gap filling sequences
produced by PCR or \emph{in silico} methods as outlined in the Background. \pb

@@

The first line of the scaffold file in Figure 1 begins with three dashes to
indicate the start of a YAML-formatted document. The first entry (highlighted
in green) begins with a dash character `-' to denote an entry in the YAML list.
This is a requirement of the YAML format: each entry begins with a dash line.
The next line is indented by two spaces where whitespace is used to group
similar attributes together. The ``sequence'' tag indicates that this entry
corresponds to a sequence in the FASTA file and the following line indicates
the name of this sequence using the ``source'' tag. The first word of the FASTA
header is used to identify which sequence is selected from the file. Together
these three lines describe the first entry in the scaffold as a simple sequence
using the sequence named `sequence1'. On the right hand side of Figure 1 this
produces the first region in the scaffold, also shown in green. \pb

@@

The last entry in the scaffold, highlighted in blue, adds the sequence named
`sequence2' to the scaffold. This entry demonstrates how this sequence may be
manipulated prior to addition to the scaffold. The `start' and `stop' tags trim
the sequence to these coordinates inclusively. The ``reverse'' tag also
instructs Scaffolder to reverse complement the sequence. In the putative
scaffold shown in Figure 1 this completes the sequence. \pb

The final entry in the scaffold uses the ``inserts'' tag to add additional
regions of sequence. These inserts are also added as a YAML list, with each
insert beginning with a dash. The first insert, shown in purple, uses similar
attributes to that of a sequence entry; the reverse, start and stop tags are
used to trim and reverse complement the insert. Similarly the `source' tag
identifies the corresponding FASTA sequence as `insert1'. The ``open'' and
``close'' tags are specific to inserts and determine where the insert is added
in the enclosing sequence. The region of the sequence inside these coordinates
is inclusively replaced by the specified insert sequence. This is visualised in
the putative scaffold in Figure 1 by the black lines bisecting the blue
sequence. \pb

The next insert, shown in brown, is specified using only the `open' tag. This
illustrates that only one of either `open' or `close' tags is required when
adding an insert sequence. If only one of the `open' or `close' tags is used
the corresponding opposing `open'/`close' coordinate is calculated from the
length of the insert FASTA sequence. This allows inserts to bridge into, and
partially fill gap regions, without requiring an end coordinate position. \pb

@@

Scaffolding an incomplete genome assembly requires joining contigs and
additional gap-filling sequences using a combination of computational and
laboratory methods. The process of manually editing a scaffold is inherently
hard to reproduce and introduces irreproducible edits and/or human error. In
respect to this the aims of the Scaffolder software are twofold: 1) to provide
software that is easy to install and simplifies the task of genome finishing;
and 2) to facilitate reproducibility in the scaffolding and finishing stage of
a genome project. \pb

Scaffolder uses a minimal and compact syntax to describe how a genome scaffold
sequence should be generated. This syntax is simple to write and edit whilst
being succinct and readable. \pb

AGP is a similar format for describing scaffolds. This format can be used to
describe contig order and N-filled gap regions in a scaffold. The advantages of
the AGP format is that each contig entry is defined on a single line which
allows searching the scaffold using UNIX line-based tools. The Scaffolder
format in comparison in contrast is written in the standardised YAML format and
therefore accessible to the many languages which provide parsers to this
format. The Scaffolder format is provided with a tool explicitly to produce the
FASTA sequence of scaffold specified by in the file. The Scaffolder format
further provides functions for trimming and replacing regions of sequence using
inserts. \pb

Constructing a genome using by specifying the scaffold organisation in text
file makes generating a scaffold super sequence both reproducible and
deterministic for the same file and set of FASTA sequences. In comparison,
joining large nucleotide sequences by hand cannot be reliably reproduced. The
scaffold file furthermore provides a human readable description of how the
scaffold is constructed. Configuring the final sequence in the scaffold file
means the build is easier to edit once already constructed. \pb

>>>>

The sequencing process generates reads from either strand of the DNA helix and
the resulting contigs constructed from these reads may represent either DNA
strand. Orientating all contigs to point in the same direction requires reverse
complementing sequences where necessary. In the case of archaeal and bacterial
genomes this orientation will be to the 5'~$\rightarrow$~3' direction following
the direction of genome replication.

@@

During the scaffolding process, closing and filling gaps between contigs
completes and improves the genome scaffold. Closing gaps may require returning
to the laboratory to perform additional sequencing or using computational
methods to estimate the unknown sequence. This additional sequence is used to
replace the gap between two contigs, joining them into a single sequence. Once
all contigs have been joined and gaps in a scaffold closed, the genome may be
considered finished.

@@

The process of finishing a genome scaffold uses wet laboratory methods,
\emph{in silico} methods, or a combination of both. An example of a
computational method might use the paired-read data from the sequencing stage.
The occurrence of paired reads in separate contigs can be used to estimate
probabilistically the order and distance between these contigs. Alternatively,
laboratory methods may use PCR to amplify the unknown DNA in a gap region then
use Sanger sequencing to determine the sequence of this gap. Computational
methods, using available sequencing data, are preferable as they are less
costly in laboratory time and materials compared to manual gap resolution
\cite{nagarajan2010}. Finally when the scaffold cannot be completely resolved,
\emph{in silico} software packages exist to suggest the likely primers
necessary for PCR amplifying the sequence in gap regions \cite{gordon2001}. \pb

@@

Initial sequencing data assembly may result in an incomplete genome build.
Adding further sequences from either PCR or computational methods also means
that genome scaffolding may be an on-going process. The scaffold file should
therefore be simple to update manually in addition to being computationally
tractable. This requirement was also best suited to YAML syntax which is
human-readable and simple to edit in a standard text editor. \pb

@@

The input data for Scaffolder is nucleotide sequence in FASTA format file.
These nucleotide sequences can be of any length and for example may be
individual reads, assembled contigs or contigs which have been joined into
larger scaffolds. The case in which Scaffolder may be most useful is using the
contigs and scaffolded contigs, combined with additional gap filling sequences
produced by PCR or \emph{in silico} methods as outlined in the Background. \pb

@@

The first line of the scaffold file in Figure 1 begins with three dashes to
indicate the start of a YAML-formatted document. The first entry (highlighted
in green) begins with a dash character `-' to denote an entry in the YAML list.
This is a requirement of the YAML format: each entry begins with a dash line.
The next line is indented by two spaces where whitespace is used to group
similar attributes together. The ``sequence'' tag indicates that this entry
corresponds to a sequence in the FASTA file and the following line indicates
the name of this sequence using the ``source'' tag. The first word of the FASTA
header is used to identify which sequence is selected from the file. Together
these three lines describe the first entry in the scaffold as a simple sequence
using a FASTA entry named `sequence1'. On the right hand side of Figure 1 this
produces the first region in the scaffold, also shown in green. \pb

@@

The last entry in the scaffold, highlighted in blue, adds a FASTA entry named
`sequence2' to the scaffold. This entry demonstrates how this sequence may be
manipulated prior to addition to the scaffold. The `start' and `stop' tags trim
the sequence to these coordinates inclusively. The ``reverse'' tag also
instructs Scaffolder to reverse complement the sequence. In the putative
scaffold shown in Figure 1 this completes the sequence. \pb

This final entry in the scaffold uses the ``inserts'' tag to add additional
regions of sequence. These inserts are also added as a YAML list, with each
insert beginning with a dash. The first insert, shown in purple, uses similar
attributes to that of a sequence entry; the reverse, start and stop tags are
used to trim and reverse complement the insert. Similarly the `source' tag
identifies the corresponding FASTA sequence as `insert1'. The ``open'' and
``close'' tags are specific to inserts and determine where the insert is added
in the enclosing sequence. The region of the sequence inside these coordinates
is inclusively replaced by the specified insert sequence. This is visualised in
the putative scaffold in Figure 1 by the black lines bisecting the blue
sequence. \pb

The next insert, shown in brown, is specified using only the `open' tag. This
illustrates that only one of either the `open' or `close' tags is required when
adding an insert sequence. If only one of the `open' or `close' tags is used
the corresponding opposing `open'/`close' coordinate is calculated from the
length of the insert FASTA sequence. This allows inserts to bridge into, and
partially fill gap regions, without requiring an end coordinate position. \pb

@@

Scaffolding an incomplete genome assembly requires joining contigs and
additional gap-filling sequences using a combination of computational and
laboratory methods. The process of manually editing a scaffold is inherently
hard to reproduce and introduces irreproducible edits and/or human error. In
respect to this the aims of the Scaffolder software are twofold: 1) to provide
software that is easy to install and simplifies the task of genome finishing;
and 2) to facilitate reproducibility in the scaffolding and finishing stage of
a genome project. Scaffolder uses a minimal and compact syntax to describe how
a genome scaffold sequence should be generated. This syntax is simple to write
and edit whilst being succinct and readable. \pb

AGP is a similar format for describing scaffolds. This format can be used to
describe contig order and N-filled gap regions in a scaffold. The advantage of
the AGP format is that each contig entry is defined on a single line which
allows searching the scaffold using UNIX line-based tools. The Scaffolder
format in contrast is written in the standardised YAML format and therefore
accessible to the many languages which provide parsers to this format. The
Scaffolder format is provided with a tool explicitly to produce the FASTA
sequence of scaffold specified by the file. The Scaffolder format further
provides functions for trimming and replacing regions of sequence using
inserts. \pb

Constructing a genome by specifying the scaffold organisation in text file
makes generating a scaffold super sequence both reproducible and deterministic
for the same file and set of FASTA sequences. In comparison, joining large
nucleotide sequences by hand cannot be reliably reproduced. The scaffold file
furthermore provides a human readable description of how the scaffold is
constructed. Configuring the final sequence in the scaffold file means the
build is easier to edit once already constructed. \pb
